Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-05-19 paper.tex
Reasonable version of the intro

[PARA 4: we present sc2ts based on these recent advances. Summarise
results, with forward refs to Results sections.]

% %% ARGs are good for *representing* viral genealogies

% %% ARG inference is hard

% Inferring recombinant phylogenies has been a long-standing computational challenge.
% However, approaches to infer ARGs now exist that can scale to tens of
% thousands of human genomes and beyond
% \citep{Speidel2019-yh,Kelleher2019-ba,Zhang2021-lf,Schaefer2021-yg}, and which
% deal with levels of repeated recombination far in excess of those seen in viral
% phylogenies. For example, using over 90 million variable sites
% \citet{Wohns2022-th} reconstructed a unified whole-genome genealogy of thousands
% of modern and ancient human genomes, accounting for millions of recombination
% events (over 5 million ``local trees''). To achieve this, \cite{Wohns2022-th}
% took advantage of a highly efficient Hidden Markov Model (HMM) implementation
% of the Li and Stephens (LS) model \citep{Li2003-ib} to infer copying paths between
% genomes \citep{Kelleher2019-ba}.

% % What is new here?

% Here, we present a preliminary version of a novel scalable methodology
% (\texttt{sc2ts}, optionally pronounced ``scoots'') that builds on the advances above
% and additionally takes advantage of known collection dates to infer ARGs from millions
% of full-length SARS-CoV-2 genomes. Using publicly available SARS-CoV-2 genome sequences,
% we demonstrate that \texttt{sc2ts} can be used to simultaneously (1) detect putative SARS-CoV-2
% recombinants; (2) infer the phylogenetic relationships of the major SARS-CoV-2 lineages;
% (3) reconstruct the mutational spectrum of SARS-CoV-2; and (4) provide a compact record
% of evolutionary information about SARS-CoV-2 that can be efficiently accessed using
% \texttt{tskit} (tree sequence toolkit) software library \citep{Kelleher2018-xc}.
% \texttt{Sc2ts} is a promising method to automate detection of SARS-CoV-2 recombinants
% and to jointly model recombinational and mutational processes shaping the genomic
% diversity of SARS-CoV-2 at scale.

\section{Results}
\subsection{Inferred ARGs}
For this preliminary study our goals are to illustrate the utility of
\texttt{sc2ts}, and to closely study the properties of the
inferred ARGs to inform subsequent development. Rather
than attempting to be comprehensive we chose to work with a
representative subset of the available data, limited to
inferences that could be performed on a single server in
a few weeks (see below for further details on timings and
computer hardware used). The cut-off dates for sampling were chosen
arbitrarily.
We inferred two ARGs, which we refer to as the
``Wide'' and ``Long'' ARGs throughout.
The Wide ARG is densely sampled but time-limited and
includes all 1.27 million sample genomes  with a
maximum sampling delay of 30 days
(see Section~\ref{sec:filtering_time_travellers})
up to June 30, 2021.
The Long ARG is more sparsely sampled but
includes data for an additional year (June 30, 2022).
For the long ARG, we use at most 1,000 randomly subsampled
genomes per day from those meeting the maximum sampling delay criterion.

@@

It is important to note that the exact endpoints of intervals can be
somewhat arbitrary, since they are defined by exact comparisons
of a recombinant's parents. The presence of problematic sites that are enriched
for errors will therefore tend to artificially truncate breakpoint intervals
at these sites.
It is therefore helpful to compare the number of interecting intervals
with the number of mutations per site in Figure~\ref{fig:breakpoint-distribution}.
For example, position 28,271 (located in the ORF8-N intergenic
region) has the largest number of mutatations and corresponds to the
left endpoint of 9 intervals and the right of 57.
The 7,572 mutations  (including 5,782 insertions and 1,605 deletions)
at this site are a likely indicator that this
homopolymeric region is prone to sequencing errors
and should be included in the list of ``problematic sites'' that
are excluded from analysis (Section~\ref{sec:data_preprocessing}).
% There can be good biological reasons for this too though
However, hypermutable sites such as 27,972 \citep{Jungreis2021-dh}
have the same effect of truncating intervals. At this site we have
770 mutations (including 425 C$>$T, 314 T$>$C) with potential signals
of polymorphism maintained by selection [FIXME!!], and 58 intervals
with endpoints within one base of this site.

@@

Sc2ts is a method for inferring ARGs from densely sampled pandemic-scale data
in real time, in which recombination occurs at a low but significant rate.
The essential idea is to incrementally update the ARG each day
with the sequences collected on that day. These updates are performed by
first finding likely ``copying paths'' for each sequence in the daily
batch to the current ARG, using the ARG-based implementation of
the Li and Stephens model~\citep{Li2003-ib} introduced
in \texttt{tsinfer}~\citep{Kelleher2019-ba}.
These copying paths
will mostly consist of a new sample sequence copying from a node in the ARG
with a small number of mutations, and often many samples
from a daily batch will copy from the same ARG node. The second
step is then to ``resolve'' these implied polytomies by using
standard tree-building techniques (Figure~\ref{fig:overview_sc2ts}C).
This greedy update strategy inevitably lead to unparsimonious
topologies, and the third step is to then increase the
overall parsimony of the inferred ARG by making some simple topological
updates (Figure~\ref{fig:overview_sc2ts}C and D).
The operations are carried out efficiently using the well-established
\texttt{tskit}
library~\citep{Kelleher2018-xc,Ralph2020-efficiently,Tskit2023-tskit},
and the resulting structure can be conveniently and interactively
analysed using the many features provided by \texttt{tskit}.

@@

The \texttt{tskit} library~\citep{Tskit2023-tskit} is a liberally
licensed open source toolkit that provides a comprehensive suite
of tools for working with ARGs. Based on core functionality written
in C, it provides interfaces in C, Python and Rust. The Python interface
is based on NumPy~\citep{Harris2020-array}, and provides a convenient
platform for interactive analysis of large-scale data using, for
example, Jupyter notebooks~\citep{Kluyver2016-jupyter} and taking
advantage of the analysis tools in the burgeoning PyData ecosystem.
Tskit is mature software, widely used in population genetics, and
has been incorporated into several downstream
applications~\citep[e.g.][]{Haller2019-slim,Speidel2019-yh,
Terasaki2021-geonomics,
Fan2022-genealogical,Korfmann2022-weak,
Mahmoudi2022-bayesian,Petr2022-slendr,Rasmussen2022-espalier}.
This ecosystem for storing and manipulating ARGs can
generally be used to efficiently record and analyse SARS-CoV-2 genealogies reconstructed
using other methods, not only the new approach that we describe here.

>>>>

% [PARA 4: we present sc2ts based on these recent advances. Summarise
% results, with forward refs to Results sections.]
Here we present a preliminary version of \texttt{sc2ts}, a novel method
for inferring ARGs for SARS-CoV-2 at pandemic scale in real-time.
Building on the open-source \texttt{tskit} library, the method infers
a joint model of mutation, genetic inheritance and recombination, which
can be conveniently and efficiently analysed using standard data science
tools. Inference is based on incrementally adding batches of sequences
based on their collection dates, and proceeds in three phases. Firstly,
likely attachment nodes in the current ARG
(or paths, if inferred to be a recombinant) are
chosen for each of the samples in the daily batch using the
Li and Stephen's model (Section~\ref{sec:ls}). Then, since many samples
in a daily batch can share an attachment path,
we infer trees for each of the clusters separately using standard
methods (Section~\ref{sec:sample-cluster-tree-inference}).
Finally, we attach the trees for these sample-clusters to the current
ARG, and apply some parsimony heuristics to address issues
introduced by the inherent greediness of the strategy
(Section~\ref{sec:parsimony-heuristics}).
Using the current preliminary version of \texttt{sc2ts} we infer two
large ARGs (1,265,685 and 657,239 samples, respectively), and study
the properties of these ARGs to illustrate the power of the method
and to inform subsequent development. We find that these
ARGs accurately reflect known patterns of phylogenetic relationships
between samples (Section~\ref{sec:backbone_phylogeny}),
and mutational spectra (Section~\ref{sec:mutation_spectrum})
while simultaneously identifying the majority of known recombinant
lineages (Sections~\ref{sec:jackson_recombs} and \ref{sec:pango_x_lineages}),
with an unprecedented degree of precision in the
genomic location of recombination breakpoints
(Section~\ref{sec:breakpoint_intervals})
and the relationship between parental sequences
(Section~\ref{sec:parent_divergence}).
We hope that these benefits of an accurate joint model of
mutation, genetic inheritance and recombination, will generate community
interest and development of the \texttt{sc2ts} method, and
more generally in applying the efficient and mature software
of the \texttt{tskit} ecosystem to pandemic-scale SARS-CoV-2 data.

\section{Results}
\subsection{Inferred ARGs}
The goals of this preliminary study are to illustrate the utility of
\texttt{sc2ts}, and to closely study the properties of the
inferred ARGs to inform subsequent development. Rather
than attempting to be comprehensive we chose to work with a
representative subset of the available data, limited to
inferences that could be performed on a single server in
a few weeks (see below for further details on timings and
computer hardware used). The cut-off dates for sampling were chosen
arbitrarily.
We inferred two ARGs, which we refer to as the
``Wide'' and ``Long'' ARGs throughout.
The Wide ARG is densely sampled but time-limited and
includes all 1.27 million sample genomes  with a
maximum sampling delay of 30 days
(see Section~\ref{sec:filtering_time_travellers})
up to June 30, 2021.
The Long ARG is more sparsely sampled but
includes data for an additional year (June 30, 2022).
For the long ARG, we use at most 1,000 randomly subsampled
genomes per day from those meeting the maximum sampling delay criterion.

@@

It is important to note that the exact endpoints of intervals can be
somewhat arbitrary, since they are defined by exact comparisons
of a recombinant's parents. The presence of problematic sites that are enriched
for errors will therefore tend to artificially truncate breakpoint intervals
at these sites.
It is therefore helpful to compare the number of interecting intervals
with the number of mutations per site in Figure~\ref{fig:breakpoint-distribution}.
For example, position 28,271 (located in the ORF8-N intergenic
region) has the largest number of mutations and corresponds to the
left endpoint of 9 intervals and the right of 57.
The 7,572 mutations  (including 5,782 insertions and 1,605 deletions)
at this site are a likely indicator that this
homopolymeric region is prone to sequencing errors
and should be included in the list of ``problematic sites'' that
are excluded from analysis (Section~\ref{sec:data_preprocessing}).
% There can be good biological reasons for this too though
However, hypermutable sites such as 27,972 \citep{Jungreis2021-dh}
have the same effect of truncating intervals. At this site we have
770 mutations (including 425 C$>$T, 314 T$>$C) with potential signals
of polymorphism maintained by selection [FIXME!!], and 58 intervals
with endpoints within one base of this site.

@@

Sc2ts (pronounced ``scoots'', optionally) is a method for inferring
ARGs from densely sampled pandemic-scale data
in real time, in which recombination occurs at a low but significant rate.
The essential idea is to incrementally update the ARG each day
with the sequences collected on that day. These updates are performed by
first finding likely ``copying paths'' for each sequence in the daily
batch to the current ARG, using the ARG-based implementation of
the Li and Stephens model~\citep{Li2003-ib} introduced
in \texttt{tsinfer}~\citep{Kelleher2019-ba}.
These copying paths
will mostly consist of a new sample sequence copying from a node in the ARG
with a small number of mutations, and often many samples
from a daily batch will copy from the same ARG node. The second
step is then to ``resolve'' these implied polytomies by using
standard tree-building techniques (Figure~\ref{fig:overview_sc2ts}C).
This greedy update strategy inevitably lead to unparsimonious
topologies, and the third step is to then increase the
overall parsimony of the inferred ARG by making some simple topological
updates (Figure~\ref{fig:overview_sc2ts}C and D).
The operations are carried out efficiently using the well-established
\texttt{tskit}
library~\citep{Kelleher2018-xc,Ralph2020-efficiently,Tskit2023-tskit},
and the resulting structure can be conveniently and interactively
analysed using the many features provided by \texttt{tskit}.

@@

The \texttt{tskit} library~\citep{Tskit2023-tskit} is a liberally
licensed open source toolkit that provides a comprehensive suite
of tools for working with ARGs. Based on core functionality written
in C, it provides interfaces in C, Python and Rust. The Python interface
is based on NumPy~\citep{Harris2020-array}, and provides a convenient
platform for interactive analysis of large-scale data using, for
example, Jupyter notebooks~\citep{Kluyver2016-jupyter} and taking
advantage of the analysis tools in the burgeoning PyData ecosystem.
(It is possible to access the toolkit from R via the \texttt{reticulate}
package, and the \texttt{slendr} library~\citep{Petr2022-slendr}
also provides some native R support. A full R interface would be a
valuable addition to the ecosystem.)
Tskit is mature software, widely used in population genetics, and
has been incorporated into several downstream
applications~\citep[e.g.][]{Haller2019-slim,Speidel2019-yh,
Terasaki2021-geonomics,
Fan2022-genealogical,Korfmann2022-weak,
Mahmoudi2022-bayesian,Petr2022-slendr,Rasmussen2022-espalier}.
It is important to note that
this ecosystem for storing and manipulating ARGs can
generally be used to efficiently record and analyse SARS-CoV-2 genealogies reconstructed
using other methods, not only the new approach that we describe here.
